Universal algebra and the \texorpdfstring{$ \lambda $}{lambda}-calculus

### 2024-06-07 report/src/algebraic-structures.tex
Set up for the final form of the fundamental theorem

@@

\begin{remark}
  Note that for an algebraic theory $ T $, the $ T_n $ are all algebras for $ T $, with the action given by $ \bullet $.
\end{remark}

\begin{definition}[Pullback of algebras]\label{def:algebra-pullback}
  If we have a morphism of algebraic theories $ f: T^\prime \to T $, we have a functor $ \Alg T \to \Alg T^\prime $. It endows $ T^\prime $-algebras with an action from $ T $ given by $ g \bullet_{T^\prime} a = f(g) \bullet_T a $. Then $ T^\prime $-algebra morphisms commute with this $ T $-action, so we indeed have a functor.
\end{definition}

@@

@@

\begin{lemma}\label{lem:presheaf-Yoneda}
  Given an algebraic theory $ T $ and a $ T $-presheaf $ Q $, we have for all $ n $ a bijection of sets
  \[ \varphi: \Pshf T(T^n, Q) \cong Q_n. \]
\end{lemma}
\begin{proof}
  For $ f: \Pshf T(T^n, Q) $, take $ \varphi(f) = f_n(x_1, \dots, x_n) $.

  Conversely, for all $ q: Q_n $ and all $ t_1, \dots, t_n: T_m^n $ take
  \[ \varphi^{-1}(q)_m(t_1, \dots, t_n) = q \bullet t. \]
\end{proof}

>>>>

\begin{remark}
  \TODO Say something about universal algebra
  For equivalent definitions of different kinds, see Chapter \ref{ch:alternative-definitions}.
\end{remark}

@@

\begin{remark}
  Note that for an algebraic theory $ T $, the $ T_n $ are algebras for $ T $, with the action given by $ \bullet $.
\end{remark}

\begin{definition}\label{def:algebra-pullback}
  If we have a morphism of algebraic theories $ f: T^\prime \to T $, we have a \textit{pullback functor of algebras}\index{pullback functor!of algebras} $ f^* : \Alg T \to \Alg {T^\prime} $. It endows $ T $-algebras with an action from $ T^\prime $ given by $ g \bullet_{T^\prime} a = f(g) \bullet_T a $. Then $ T $-algebra morphisms commute with this $ T^\prime $-action, so we indeed have a functor.
\end{definition}

\begin{remark}
  Note that by Lemma \ref{lem:lawvere-algebra}, algebras for $ T $ are equivalent to finite-product-preserving functors from its Lawvere theory to $ \SET $. Then $ f: T^\prime \to T $ corresponds to a functor on the Lawvere theories $ \mathbf L_f: \mathbf L_{T^\prime} \to \mathbf L_T $, and $ f^*: \Alg T \to \Alg{T^\prime} $ corresponds to precomposition with $ \mathbf L_f $:
  \begin{center}
    \begin{tikzcd}
      \mathbf L_{T^\prime} \arrow[r, "{\mathbf L_f}"] \arrow[rd, dashed] & \mathbf L_T \arrow[d]\\
      & \SET
    \end{tikzcd}
  \end{center}
\end{remark}

Actually, we can recover some information about the algebraic theory morphism from this pullback functor. For example,
\begin{lemma}\label{lem:algebra-pullback-equivalence}
  If $ F^* $ is an equivalence of categories, $ F $ is an isomorphism.
\end{lemma}
\begin{proof}
  \TODO
\end{proof}

@@

\begin{remark}
  Recall from \ref{lem:lawvere-algebra} that $ T $-algebras are equivalent to finite-product-preserving $ \SET $-valued functors on the Lawvere theory $ \mathbf L $ associated to $ T $. Now, recall from (\TODO) that we can embed any category inside its category of $ \SET $-valued functors. This embeds $ n: \mathbf L $ as the theory algebra $ T_n $. Then the Yoneda Lemma gives a bijection
  \[ \Alg T(T_n, A) \cong A^n = \SET(\{ 1, \dots, n \}, A) \]
  natural in $ n $ and $ A $. Explicitly, it sends $ F: \Alg T(T_n, A) $ to $ i \mapsto F(x_i) $ and $ (a_i)_i : A^n $ to $ f \mapsto f \bullet a $. The natural equivalence immediately shows that $ T_n $ is the free $ T $-algebra on $ n $ elements.
\end{remark}

@@

\begin{remark}\label{rem:presheaf-Yoneda}
  Recall from \ref{lem:lawvere-presheaf} that $ T $-presheaves are equivalent to presheaves on the Lawvere theory $ \mathbf L $ associated to $ T $. Now, recall from Definition \ref{def:Yoneda-embedding} that we can embed any category inside its own category of presheaves. This embeds $ n: \mathbf L $ as the power $ T^n = (T_m^n)_m $ of the theory presheaf. Then the Yoneda Lemma gives a bijection
  \[ \Pshf T(T^n, Q) \cong Q_n \]
  natural in $ n $ and $ Q $. Explicitly, it sends $ f: \Pshf T(T^n, Q) $ to $ f_n(x_1, \dots, x_n) $ and $ q: Q_n $ to $ (t_i)_i \mapsto q \bullet t $.
\end{remark}

### 2024-06-07 report/src/alternative-definitions.tex
Define the theory of extensions
Set up for the final form of the fundamental theorem

\chapter{Alternative definitions}
The literature, there are many different but equivalent definitions, carrying many different names, for the objects that called `algebraic theories' in Section \ref{sec:algebraic-theories}. Also, many of these different names are attached to differently defined objects in different sources. This section will showcase some of the various definitions.

@@

\begin{remark}
  The definition of algebraic theory that Hyland gives is closest to that of an abstract clone. However, instead of a sequence of sets $ (T_n)_n $, he requires a functor $ T: F \to \SET $ (with $ F \subseteq \FinSET $ the skeleton category of finite sets $ F_0 = \{ \llbracket 0 \rrbracket, \llbracket 1 \rrbracket, \dots \} $), and he requires $ \bullet: T_m \times T_n^m \to T_n $ to be dinatural in $ n $ and natural in $ m $.

@@

\begin{lemma}
  Let $ C $ be an abstract clone, and $ L $ be its associated Lawvere theory by the equivalence given above. A $ C $-algebra is equivalent to an algebra for $ L $.
\end{lemma}
\begin{proof}
  Let $ A $ be a $ C $-algebra. We will construct a functor $ F: L \to \SET $ as follows: We take $ F(n) = A^n $. We define the action on morphisms as
  \[ F(f)(a) = (f_i \bullet a)_i \]
  for $ f: L(m, n) = L(m, 1)^n $ and $ a: A^m $.

>>>>

\chapter{Alternative definitions}\label{ch:alternative-definitions}
The literature, there are many different but equivalent definitions, carrying many different names, for the objects that called `algebraic theories' in Section \ref{sec:algebraic-theories}. Also, many of these different names are attached to differently defined objects in different sources. This section will showcase some of the various definitions.

@@

\begin{remark}\label{rem:abstract-clone-functor}
  The definition of algebraic theory that Hyland gives is closest to that of an abstract clone. However, instead of a sequence of sets $ (T_n)_n $, he requires a functor $ T: F \to \SET $ (with $ F \subseteq \FinSET $ the skeleton category of finite sets $ F_0 = \{ \llbracket 0 \rrbracket, \llbracket 1 \rrbracket, \dots \} $), and he requires $ \bullet: T_m \times T_n^m \to T_n $ to be dinatural in $ n $ and natural in $ m $.

@@

\begin{lemma}\label{lem:lawvere-algebra}
  Let $ C $ be an abstract clone, and $ L $ be its associated Lawvere theory by the equivalence given above. A $ C $-algebra is equivalent to an algebra for $ L $.
\end{lemma}
\begin{proof}
  Let $ A $ be a $ C $-algebra. We will construct a functor $ F: L \to \SET $ as follows: We take $ F(n) = A^n $. We define the action on morphisms as
  \[ F(f)(a) = (f_i \bullet a)_i \]
  for $ f: L(m, n) = L(m, 1)^n $ and $ a: A^m $.

### 2024-06-07 report/src/category-theory.tex
Set up for the final form of the fundamental theorem

@@

\section{Yoneda}
We can embed a category $ C $ fully faithfully into the functor category $ P C = [\op C, \SET] $ as follows (see \autocite{Kashiwara}, Section 1.4):
\begin{definition}
  The \iindex{Yoneda embedding} $ Y : C \hookrightarrow P C $ is given on objects by $ Y(c) = C(-, c) $:
  \[ Y(c)(d) = C(d, c) \quad \text{and} \quad Y(c)(f)(g) = f \cdot g \]
  for $ d: C $, $ f: C(d, d^\prime) $ and $ g: C(d^\prime, c) $. It sends a morphism $ f: C(c, c^\prime) $ to the natural transformation $ Y(f): C(-, c) \Rightarrow C(-, c^\prime) $ given by
  \[ Y(f)(d)(g) = g \cdot f \]
  for $ d: C $ and $ g: C(d, c) $.
\end{definition}

@@

\begin{lemma}
  The Yoneda embedding functor preserves exponential objects (\autocite{stackexchange:yoneda-exponentials}).
\end{lemma}
\begin{proof}

\end{proof}

>>>>

Alternatively, we can characterize universal arrows by their action on hom-sets:
\begin{lemma}
  Let $ F: C \to D $ be a functor and $ d: D $ an object. An object $ c: C $ and an arrow $ f: D(d, F(c)) $ form a universal arrow from $ d $ to $ F $ if and only if the function
  \[ (g \mapsto f \cdot F(g)) : C(c, x) \cong D(d, F(x)) \]
  is a bijection.

  Conversely, for all $ c: C $ and $ d: D $, every bijection
  \[ C(c, x) \cong D(d, F(x)) \]
  that is natural in $ x $ arises in this way from some universal arrow $ f: D(d, F(c)) $.
\end{lemma}
\begin{proof}
  See \autocite{MacLane}, Chapter III.2, Proposition 1.
\end{proof}

@@

\subsection{Forgetful functors and free objects}

In mathematics, we often deal with objects that are `based on' other objects. For example, a ring is a set with some additional structure. Often, this is a relation between the respective categories (for example, in the case of a displayed category, see Section \ref{sec:displayed-categories}), and such a relation gives rise to a \iindex{forgetful functor}, that `forgets' about the additional structure. In the examples of rings and sets, the forgetful functor sends a ring to its underlying set, and a ring morphism to the function between the sets. However, note that there is no formal definition of forgetful functors. The name is more of a way to talk about the perceived relation between the categories.

\begin{definition}
  Given a forgetful functor $ F: C \to D $, we define the \textit{free functor}\index{free!functor} associated to $ F $ to be the left adjoint to $ F $, if it exists.
\end{definition}

\begin{example}
  Consider the forgetful functor from the category of commutative rings to the category of sets, sending a ring to its underlying set. This has a left adjoint, sending the set $ \{ 1, 2, \dots, n \} $ to the polynomial ring $ \mathbb Z[X_1, \dots, X_n] $, and more generally, sending $ S $ to the polynomial ring $ \mathbb Z[X_s]_{s : S} $. This ring is then called `the free commutative ring on $ S $'. If $ S $ is finite of size $ n $, the ring is also called `the free commutative ring on $ n $ generators'.

  The free functor sends a function $ f: S \to T $ to the ring morphism $ \mathbb Z[X_s]_{s: S} \to \mathbb Z[X_t]_{t: T} $ that sends $ X_s $ to $ X_{f(s)} $.

  The natural bijection
  \[ \mathbf{Rng}(\mathbb Z[X_s]_{s : S}, R) \cong \SET(S, R) \]
  then sends $ f: \mathbf{Rng}(\mathbb Z[X_s]_{s : S}, R) $ to $ s \mapsto f(X_s) $ and $ g: \SET(S, R) $ to the morphism that sends $ X_s $ to $ g(s) $.
\end{example}

However, sometimes, we have a forgetful functor $ F: C \to D $, but we cannot give a free functor on the entire category $ D $. In such a case, we might still talk about free `objects':
\begin{definition}
  Let $ F: C \to D $ be a forgetful functor. Given $ d: D $, the \textit{free object}\index{free!object} on $ d $ is a universal arrow $ (c, f) $ from $ d $ to $ F $.
\end{definition}

\begin{remark}
  By \autocite{MacLane}, Chapter IV.1, Theorem 2 (ii), if we have a free object on every $ d: D $, we can piece these together to get a free functor associated to $ F $.
\end{remark}

\section{Yoneda}
We can embed a category $ C $ fully faithfully into the functor category $ P C = [\op C, \SET] $ as follows (see \autocite{Kashiwara}, Section 1.4):
\begin{definition}\label{def:Yoneda-embedding}
  The \iindex{Yoneda embedding} $ Y : C \hookrightarrow P C $ is given on objects by $ Y(c) = C(-, c) $:
  \[ Y(c)(d) = C(d, c) \quad \text{and} \quad Y(c)(f)(g) = f \cdot g \]
  for $ d: C $, $ f: C(d, d^\prime) $ and $ g: C(d^\prime, c) $. It sends a morphism $ f: C(c, c^\prime) $ to the natural transformation $ Y(f): C(-, c) \Rightarrow C(-, c^\prime) $ given by
  \[ Y(f)(d)(g) = g \cdot f \]
  for $ d: C $ and $ g: C(d, c) $.
\end{definition}

@@

\begin{lemma}
  The Yoneda embedding functor is \iindex{cartesian}. That is, not only does it preserve binary products, but it also preserves exponential objects (\autocite{stackexchange:yoneda-exponentials}).
\end{lemma}
\begin{proof}
  \TODO
\end{proof}

### 2024-06-07 report/src/paper.tex
Set up for the final form of the fundamental theorem

  Lemma \ref{lem:presheaf-Yoneda} gives a sequence of bijections $ \varphi_n : \Pshf L(L^n, L) \cong L_n $ for all $ n $, sending $ F: \Pshf L(L^n, L) $ to $ F(x_1, \dots, x_n) $, and conversely sending $ s: L_n $ to $ ((t_1, \dots, t_n) \mapsto s \bullet (t_1, \dots, t_n)) $. It considers $ \lambda $-terms in $ n $ variables as $ n $-ary functions on the $ \lambda $-calculus. Therefore, it should come as no surprise that $ \varphi $ preserves the $ x_i $, $ \bullet $, $ \rho $ and $ \lambda $, which makes it into an isomorphism of $ \lambda $-theories and this concludes the proof.
\end{proof}

@@

Let $ \llbracket n \rrbracket $ denote the finite set $ \{ 1, 2, \dots, n \} $. Let \index{$ \F $}$ \F \subseteq \FinSET $ be the full subcategory with $ \F_0 = \{ \llbracket 0 \rrbracket, \llbracket 1 \rrbracket, \dots \} $.

@@

  Explicitly, we can express the coproduct of algebras, and especially its set, as the following coend (\autocite{Hyland}, Proposition 2.5) (see also Section \ref{sec:coends} for more on coends)
  \[ A + B = \int^{(m, n) : \F \times \F} A^m \times B^n \times T_{m^\prime + n^\prime}, \]
  considering $ T $ as a functor $ T : \F \to \SET $ as a functor (see Remark \ref{rem:abstract-clone-functor}). Note that we do not need the exact definition of $ A + B $ for the rest of this section. Nonetheless, it is interesting to see how it is defined and why this definition works.

@@

  The relations then give `associativity' of this substitution and also assure that reordering or duplicating $ x_i $ and their corresponding $ a_i $ and $ b_j $ do not yield different elements: associating the images on the left and right of
  \[ A^{m^\prime} \times B^{n^\prime} \times T_{m^\prime + n^\prime} \leftarrow A^{m^\prime} \times B^{n^\prime} \times T_{m + n} \rightarrow A^m \times B^n \times T_{m + n} \]
  gives
  \[ (t \bullet ((x_{f(i)})_i + (x_{m + g(j)})_j)) \bullet a + b = t \bullet ((a_{f(i)})_i + (b_{g(j)})_j) \]
  for $ f: \llbracket m \rrbracket \to \llbracket m^\prime \rrbracket $, $ g: \llbracket n \rrbracket \to \llbracket n^\prime \rrbracket $, $ t : T_{m + n} $, $ a : A^{m^\prime} $ and $ b : B^{n^\prime} $.

  Then it becomes clear what the action of $ T $ will be on $ A + B $, although the precise definition has quite a combinatorial nature. For $ s : T_l $, $ t_i : T_{m_i + n_i} $, $ a_i : A^{m_i} $ and $ b_i : B^{n_i} $, define the disjoint embeddings
  \[ d_i : \left\llbracket m_i + n_i \right\rrbracket \cong \left\llbracket m_i \right\rrbracket \sqcup \left\llbracket n_i \right\rrbracket \hookrightarrow \left\llbracket \sum_j m_j \right\rrbracket \sqcup \left\llbracket \sum_j n_j \right\rrbracket \cong \left\llbracket \sum_j m_j + \sum_j n_j \right\rrbracket, \]
  which we will use to make sure that the $ x_j $ in the different $ t_i $ are mapped to distinct variables. Then we can define
  \[ s \bullet (t_i \bullet a_i + b_i) = (s \bullet (t_i \bullet (x_{d_i(j)})_j)_i) \bullet (a_1 + b_1 + \dots + a_l + b_l). \]
  More formally (using the coend injections $ A^{\sum_k m_k} \times B^{\sum_k n_k} \times T_{\sum_k m_k + \sum_k n_k} \to A + B $), this gives functions
  \[ T_l \to (A^{m_1} \times B^{n_1} \times T_{m_1 + n_1} \to (\dots \to (A^{m_l} \times B^{n_l} \times T_{m_l + n_l} \to A + B) \dots )), \]
  commuting with the relations between the different $ (A^m \times B^n \times T_{m + n}) $, which, by repeatedly using the universal property of the coend, then correspond to functions
  \[ T_l \to (A + B \to (\dots \to (A + B \to A + B) \dots )), \]
  or, equivalently, a function
  \[ T_l \times (A + B)^l \to A + B. \]

@@

\begin{definition}[Theory of extensions]
  Let $ T $ be an algebraic theory and $ A $ a $ T $-algebra. We can define an algebraic theory $ T_A $ called `the theory of extensions of $ A $' with $ (T_A)_n = T_n + A $. The left injection of the variables $ x_i : T_n $ gives the variables.
  Now, take $ h: (T_n + A)^m $. Sending $ g: T_m $ to $ g \bullet h $ gives a $ T $-algebra morphism $ T_m \to T_n + A $. Together with the right injection morphism of $ A $ into $ T_n + A $, this gives us a $ T $-algebra morphism from the coproduct: $ T_m + A \to T_n + A $. In particular, we have a function on sets $ (T_m + A) \times (T_n + A)^m \to T_n + A $, which we will define our substitution to be.
\end{definition}

>>>>

  Remark \ref{rem:presheaf-Yoneda} gives a sequence of bijections $ \varphi_n : \Pshf L(L^n, L) \cong L_n $ for all $ n $, sending $ F: \Pshf L(L^n, L) $ to $ F(x_1, \dots, x_n) $, and conversely sending $ s: L_n $ to $ ((t_1, \dots, t_n) \mapsto s \bullet (t_1, \dots, t_n)) $. It considers $ \lambda $-terms in $ n $ variables as $ n $-ary functions on the $ \lambda $-calculus. Therefore, it should come as no surprise that $ \varphi $ preserves the $ x_i $, $ \bullet $, $ \rho $ and $ \lambda $, which makes it into an isomorphism of $ \lambda $-theories and this concludes the proof.
\end{proof}

@@

Let $ \llbracket n \rrbracket $ denote the finite set $ \{ 1, 2, \dots, n \} $. For $ T $ an algebraic theory, let $ \mathbf L $ be its corresponding Lawvere theory (Lemma \ref{lem:lawvere-clone}).

\TODO: Move the coproducts and theory of extensions to chapter 3?

@@

  Explicitly, we can express the coproduct of algebras, and especially its set, as the following coend (\autocite{Hyland}, Proposition 2.5) (see also Section \ref{sec:coends} for more on coends)
  \[ A + B = \int^{(m, n) : \mathbf L \times \mathbf L} T_{m^\prime + n^\prime} \times A^m \times B^n, \]
  considering $ A $ as a covariant functor on $ \mathbf L $ (see Lemma \ref{lem:lawvere-algebra}) and the theory presheaf $ T $ as a presheaf (see Lemma \ref{lem:lawvere-presheaf}).

  Note that we do not need the exact definition of $ A + B $ for the rest of this section. Nonetheless, it is interesting to see how it is defined and why this definition works.

@@

  However, the coend is a quotient of $ \coprod T_{m + n} \times A^m \times B^n $ along some relations. These relations then give `associativity' of this substitution $ t \bullet (a + b) $. In particular, they assure that reordering or duplicating $ x_i $ and their corresponding $ a_i $ and $ b_j $ do not yield different elements. For $ f: \mathbf L(m, m^\prime) = T_m^{m^\prime} $, $ g: \mathbf L(n, n^\prime) = T_n^{n^\prime} $, associating the images on the left and right of
  \[ T_{m^\prime + n^\prime} \times A^{m^\prime} \times B^{n^\prime} \leftarrow T_{m^\prime + n^\prime} \times A^m \times B^n \rightarrow T_{m + n} \times A^m \times B^n \]
  gives
  \[ t \bullet ((f_i \bullet a)_i + (g_j \bullet b)_j) = (t \bullet ((f_i \bullet (x_{m + n, j})_j)_i + (g_i \bullet (x_{m + n, m + j})_j)_i)) \bullet (a + b) \]
  for $ t : T_{m^\prime + n^\prime} $, $ a : A^m $ and $ b : B^n $.

  Then it becomes clear what the action of $ T $ will be on $ A + B $, although the precise definition looks a bit complex because we have to juggle a bit with the variables in the different $ T_{m + n} $. For $ s : T_l $, $ t_i : T_{m_i + n_i} $, $ a_i : A^{m_i} $ and $ b_i : B^{n_i} $, define the disjoint embeddings
  \[ d_i : \left\llbracket m_i + n_i \right\rrbracket \cong \left\llbracket m_i \right\rrbracket \sqcup \left\llbracket n_i \right\rrbracket \hookrightarrow \left\llbracket \sum_j m_j \right\rrbracket \sqcup \left\llbracket \sum_j n_j \right\rrbracket \cong \left\llbracket \sum_j m_j + \sum_j n_j \right\rrbracket, \]
  which we will use to make sure that the $ x_j $ in the different $ t_i $ are mapped to distinct variables. Then we can define
  \[ s \bullet (t_i \bullet a_i + b_i) = (s \bullet (t_i \bullet (x_{d_i(j)})_j)_i) \bullet (a_1 + b_1 + \dots + a_l + b_l). \]
  More formally (using the coend injections $ A^{\sum_k m_k} \times B^{\sum_k n_k} \times T_{\sum_k m_k + \sum_k n_k} \to A + B $), this gives functions
  \[ T_l \to (A^{m_1} \times B^{n_1} \times T_{m_1 + n_1} \to (\dots \to (A^{m_l} \times B^{n_l} \times T_{m_l + n_l} \to A + B) \dots )), \]
  commuting with the relations between the different $ (A^m \times B^n \times T_{m + n}) $, which, by repeatedly using the universal property of the coend, then correspond to functions
  \[ T_l \to (A + B \to (\dots \to (A + B \to A + B) \dots )), \]
  or, equivalently, a function
  \[ T_l \times (A + B)^l \to A + B. \]

@@

\begin{definition}[Theory of extensions]
  Let $ T $ be an algebraic theory and $ A $ a $ T $-algebra. We can define an algebraic theory $ T_A $ called the \iindex{theory of extensions} of $ A $' with $ (T_A)_n = A + T_n $. The right injection of the variables $ x_i : T_n $ gives the variables.

  For $ h: (A + T_n)^m $, sending $ g: T_m $ to $ g \bullet h $ gives a $ T $-algebra morphism $ T_m \to T_n + A $. Together with the right injection morphism of $ A $ into $ T_n + A $, this gives us a $ T $-algebra morphism from the coproduct: $ T_m + A \to T_n + A $. Doing this for every $ h: (A + T_n)^m $ gives us the substitution $ (T_m + A) \times (T_n + A)^m \to T_n + A $.

  Showing that this is indeed an algebraic theory involves invoking the universal property of the coproduct and using properties of $ T $-algebras and $ T $-algebra morphisms.
\end{definition}

\begin{remark}
  Note that for a $ T $-algebra morphism $ f: A \to B $, we have morphisms $ f: A + T_n \xrightarrow{f + \id{T_n}} B + T_n $.
\end{remark}

\begin{remark}
  Note that the right embeddings $ r_n : T_n \to A + T_n $ give an algebraic theory morphism $ (r_n)_n : T \to T_A $, so we can think of $ T $ as lying inside $ T_A $.
\end{remark}

The following result explains why we are interested in the theory of extensions:
\begin{lemma}\label{lem:extensions-algebra-coslice}
  For $ T $ an algebraic theory and $ A $ a $ T $-algebra, we have an adjoint equivalence $ \Alg{T_A} \cong (A \downarrow \Alg T) $ between algebras for $ T_A $ and the coslice category over $ A $.
\end{lemma}
\begin{proof}
  Let $ B $ be a $ T_A $-algebra. Pullback along the embedding $ (r_n)_n : T \to T_A $ gives $ (r_n)_n^*(B) : \Alg T $. Also, we have a $ T $-algebra morphism $ A \to B $ given by the composition
  \[ A \to (T_A)_0 \xrightarrow{f \mapsto f \bullet ()} B. \]

  Conversely, take $ f: \Alg T(A, B) $. For $ b: B^n $, we have a $ T $-algebra morphism $ T_n \to B $ given by $ f \mapsto f \bullet b $. This, together with $ f $, gives a morphism from the coproduct $ A + T_n \to B $, and doing this for every $ b: B^n $ gives a $ T_A $-action on $ B $ as functions $ (A + T_n) \times B^n \to B $.

  Now, showing that the function $ A \to (T_A)_0 \to B $ defined above is indeed a $ T $-algebra morphism and that the other $ B $, together with the given $ T_A $-action is indeed a $ T_A $-algebra, and furthermore showing that these extend to functors that together form an adjoint equivalence, involves checking a lot of details. One can indeed check that all of this holds, using the properties of algebraic theories, algebras, algebra morphisms and coproducts, as well as the fact that for all $ b : B^n $, $ f \mapsto f \bullet b $ is a morphism in $ \Alg T(A + T_n, (r_n)_n^*(B)) $. However, for the sake of brevity, we will omit these and point to the formalization for the details.
\end{proof}

\begin{example}
  Take $ T_n = Z[X_1, \dots, X_n] $, the polynomial rings in $ n $ variables. $ T $-algebras correspond to commutative rings, so we can call $ T $ `the theory of commutative rings'. Now, for a commutative ring $ R $, $ \Alg{T_R} $ is equivalent to the coslice category $ (R \downarrow \Alg T) $, which is the category of $ R $-algebras. Therefore, the theory of extensions $ T_R $ can be considered to be the theory of $ R $-algebras.
\end{example}

\begin{lemma}\label{lem:factor-through-extensions}
  Any algebraic theory morphism $ f: \AlgTh(S, T) $ factorizes through the embedding of $ S $ into the theory of extensions of the pullback of the algebra $ T_0 $:
  \begin{center}
    \begin{tikzcd}
      S \arrow[r] \arrow[rr, bend right, "f"'] & S_{f^*(T_0)} \arrow[r] & T
    \end{tikzcd}
  \end{center}
\end{lemma}
\begin{proof}
  For any $ n $, we have a map of $ S $-algebras $ [t \mapsto t \bullet (), f_n] : f^*(T_0) + S_n \to f^*(T_n) $, given on $ f^*(T_0) $ by $ t \mapsto t \bullet () $ and on $ S_n $ by $ f_n $. By the universal property of the coproduct, the following diagram commutes for all $ n $:
  \begin{center}
    \begin{tikzcd}[sep = large]
      S_n \arrow[r, "r_n"] \arrow[rr, bend right, "f_n"'] & f^*(T_0) + S_n \arrow[r, "{[t \mapsto t \bullet (), f_n]}"] & T
    \end{tikzcd}
  \end{center}
  which shows that $ (r_n)_n \cdot [f \mapsto f \bullet (), f_n]_n = f $.

  Now, to show that $ [f \mapsto f \bullet (), f_n]_n $ indeed constitutes an algebraic theory morphism is a bit more work. It involves using the universal property of the coproduct a couple of times, as well as showing that for all $ g : S_{f^*(T_0)}^n $,
  \[ s \mapsto ([t \mapsto t \bullet (), f_m] x) \bullet_T ([t \mapsto t \bullet (), f_n] g_i)_i \]
  is a morphism in $ \Alg S(f^*(T_0) + T_m, f^*(T_n)) $. For more details, we again point to the formalization.
\end{proof}

Now, given a $ \Lambda $-algebra $ A $, applying the above to the initial morphism $ \iota_\Lambda : \Lambda \to E(U_A) $, we get the following diagram:
\begin{center}
  \begin{tikzcd}[sep = large]
    \Lambda \arrow[r] & \Lambda_{E(U_A)_0} \arrow[r, "{[t \mapsto t \bullet (), \iota_{\Lambda_n}]_n}"] \arrow[d, "\Lambda_{\epsilon_A}"] & E(U_A)\\
    & \Lambda_A
  \end{tikzcd}
\end{center}
For the final form of the fundamental theorem, we need to show that
\[ \Lambda_{\epsilon_A}^{-1} \cdot [t \mapsto t \bullet (), \iota_{\Lambda_n}]_n = [a \mapsto \epsilon_A^{-1}(a) \bullet (), \iota_{\Lambda_n}]_n \]
is an isomorphism. By Lemma \ref{lem:factor-through-extensions}, this is equivalent to its pullback $ [a \mapsto \epsilon_A^{-1}(a) \bullet (), \iota_{\Lambda_n}]_n^* : \Alg{E(U_A)} \to \Alg{\Lambda_A} $ being an equivalence of categories.

\begin{lemma}
  $ [a \mapsto \epsilon_A^{-1}(a) \bullet (), \iota_{\Lambda_n}]_n^* $ is essentially surjective.
\end{lemma}
\begin{proof}
  Take $ B : \Alg{\Lambda_A} $. By Lemma \ref{lem:extensions-algebra-coslice} we can consider $ B $ to be a coslice $ h: \Alg{\Lambda}(A, B) $. By the equivalence of categories we established earlier, $ h $ induces a morphism $ \bar h: \AlgTh(E(U_A), E(U_B)) $. Pulling the $ E(U_B) $-algebra $ E(U_B)_0 $ back along this morphism, gives a $ E(U_A) $-algebra $ \bar h^*(E(U_B)_0) $.
\end{proof}

\begin{lemma}
  $ [a \mapsto \epsilon_A^{-1}(a) \bullet (), \iota_{\Lambda_n}]_n^* $ is fully faithful.
\end{lemma}
\begin{proof}
  \TODO
\end{proof}

\begin{lemma}
  $ \Lambda_A $ is isomorphic to $ E(U_A) $ as an algebraic theory.
\end{lemma}
\begin{proof}
  By the previous two lemmas, the pullback functor $ [a \mapsto \epsilon_A^{-1}(a) \bullet (), \iota_{\Lambda_n}]_n^* $ is a weak equivalence. Since algebra categories are univalent, this means that the pullback functor is an equivalence of categories. Then, by Lemma \ref{lem:algebra-pullback-equivalence}, $ [a \mapsto \epsilon_A^{-1}(a) \bullet (), \iota_{\Lambda_n}]_n : \AlgTh(\Lambda_A, E(U_A)) $ is an isomorphism.
\end{proof}

The final form of Hyland's representation theorem is the following:
\begin{theorem}
  The functor that sends a $ \lambda $-theory $ L $ to the $ \Lambda $-algebra $ L_0 $ and the functor that sends a $ \Lambda $-algebra $ A $ to the theory of extensions $ \Lambda_A $ form an adjoint equivalence
  \[ \LamTh \cong \Alg \Lambda. \]
\end{theorem}
